Supporting Information for\\ ``A weakly structured stem for human origins in Africa''

### 2022-12-23 supplement.tex
Update section on fixed OOA time

In our preliminary models, the unconstrained inferred split times between East
African populations and Europeans was inferred to have occurred by around
$40$ka, which is more recent than we may have expected given the
well-documented presence of modern humans outside of Africa at that time
\citep[e.g.]{Hublin2020-nk,Hajdinjak2021-jo}, as well as previous
genetics-based dating of splits between African and Eurasian populations
\citep[e.g.]{Jouganous2017-pq,Kamm2020-vn}. The unconstrained inferred split
between East and West African populations occurred about $10$ka earlier. These
recent split dates seem to be at odds with archaeological evidence for earlier
Out-of-Africa (OOA) expansion, and with the estimated date for Neanderthal
admixture into OOA populations. The parameter uncertainty due to finite genomes
does not account for this difference: in the high-likelihood models, the $95\%$
confidence intervals for the East African-European split was roughly $38 \pm
5$ka, and for the East/West African split was $45 \pm 10$ka, with the
uncertainties reflecting 2 standard deviations according to bootstrap of
genomic regions.  Thus we expect these discrepancies to result from model
misspecification in the details of the divergence between the ancestors of OOA
and east African populations. Our models suppose a single split prior to the
OOA event followed by constant symmetric migrations between eastern and OOA
populations, which is certainly an oversimplification.

Since a detailed model for the Out of Africa and Back-to-Africa was not our
main focus, we chose to keep this simplified model, but fix the east
African-Eurasian split at $50$ka \aprcomment{references of archaeological
evidence of humans in Europe/near East by around 45ka}. This also allowed us to
include Neanderthal admixture into the OOA population at the
previously-estimated $45$ka \citep{}. Fixing these parameters (i.e., east
Africa-Eurasian split at 50ka, Neanderthal admixture at 45ka) resulted in an
estimated value for the split between the ancestors of East and West African
populations at around $60$ka.  \aprcomment{here: agrees with previous estimates
    of West African-European splits from LD \citep{Ragsdale2019-nt}, which used
    different populations from the Thousand Genomes Project than here. It is
    also in close alignment with inferred divergence times from an SFS-based
    demographic inference ($\sim65$ka), which used synonymous variation from a
subset of populations used in this study -- Mende (MSL), British (GBR), and Han
Chinese (CHB) (Section or Supp fig?).} We took this value (60ka) as fixed in
most subsequent analyses. We discuss the sensitivity of our model to these
assumptions in Section~\ref{sec:change-dates}.

%\sgcomment{I feel like we should improve the discussion of relationship
%with previous estimates below, or remove.}
%
%In addition to archaeological evidence, previous genetic dating suggested 
%earlier dates for a split between European and African populations. Most of 
%these results focused on populations of recent West African ancestry. 
%
%Additionally, most previous
%estimates of the time of the OOA expansion compared Eurasian and West African
%populations. Our inferred time of the East/West African divergence aligns with
%previous estimates of the OOA event at $\approx 60$ka (themselves inferred from
%oversimplified demographic models).

@@

\begin{figure}

    \includegraphics{figures/toy_models/toy_model_combo}

    \caption{The impact of model specification on coalescence rate and inference.
    We show five simplified demographic models that differ in demographic events, 
    occurring between 200ka and 100ka, that mark the transition between a single 
    ancestral population and two daughter populations. A) A clean split 100ka, 
B) A clean split 200ka, C) population growth at 200ka followed by split 100ka, 
D) split 200ka followed by migration until 100ka, and E) split 200ka followed by 
merger 105ka and a new split 100ka. The left column shows model representation, 
the middle column shows instantaneous coalescence rates for pairs of lineages from the
same daughter population, the right column shows instantaneous coalescence rates for
pairs of lineages coming from distinct populations.  
        \label{fig:toymodels}
    }

    \end{figure}

@@

\subsection{Batch effects from merged data biases inference}\label{sec:merged-bias}

>>>>

\subsubsection{Fixing the East African and OOA divergence times}

In our preliminary models, the unconstrained inferred split times between East
African populations and Europeans was inferred to have occurred around $40$ka,
and the inferred split between East and West African populations occurred about
$10$ka earlier. These recent split dates are at odds with archaeological
evidence for an earlier Out-of-Africa (OOA) expansion
\citep[e.g.]{Hublin2020-nk,Hajdinjak2021-jo}, with the estimated date for
Neanderthal admixture into Eurasian populations \citep[e.g.][]{Reilly2022-ym},
and with previous genetics-based dating of splits between African and Eurasian
populations \citep[e.g.]{Gutenkunst2009-ff,Jouganous2017-pq,Kamm2020-vn}.  The
parameter uncertainty of our inferred divergence times (which accounts for the
non-independence of sites) does not account for these differences: in the
high-likelihood models, the $95\%$ confidence intervals for the East
African-European split was roughly $38 \pm 5$ka, and for the East/West African
split was $45 \pm 10$ka, with the uncertainties reflecting 2 standard
deviations according to bootstrap of genomic regions. We therefore expect these
discrepancies to result from model misspecification in the details of the
divergence between the ancestors of OOA and East African populations.

Our models suppose a single split leading to the OOA event followed by constant
symmetric migrations between eastern African and OOA populations, which is an
oversimplification. Previous studies that have focused on reconstructing the
OOA divergence time and subsequent migration rates typically included more
Eurasian populations and multiple epochs that allow for differing continuous
migration rates and population sizes. Since a detailed model for the OOA (and
Back-to-Africa) event was not our main focus, we chose to keep this simplified
model, but fix the east African-Eurasian split at $50$ka \citep{Hublin2020-nk}.
This also allowed us to include Neanderthal admixture into the OOA population
at $45$ka.

Fixing these parameters (i.e., East Africa-Eurasian split at 50ka, Neanderthal
admixture at 45ka) resulted in an estimated value for the split between the
ancestors of East and West African populations at around $60$ka. This timing
agrees with previous LD-based estimates of West African-Eurasian divergence
times \citep{Ragsdale2019-nt}, which used different populations than those used
in this study. A date of $60$ka is also in close agreement with inferred
divergence times from an SFS-based demographic inference (of $\sim65$ka), using
synonymous variation from a subset of populations used in this study (Supp.
Figure~\ref{fig:supp-OOA-SFS}) \aprcomment{TODO: add this figure}.
We therefore took this value ($60$ka) as fixed
in most subsequent analyses. We discuss the sensitivity of our model to these
assumptions in Section~\ref{sec:change-dates}.

@@

\begin{figure}[ht!]
    \centering
    \includegraphics{figures/toy_models/toy_model_combo}
    \caption{
        \textbf{The impact of model specification on coalescence rate and
        inference.} We show five simplified demographic models that differ in
        demographic events, occurring between 200ka and 100ka, that mark the
        transition between a single ancestral population and two daughter
        populations. A: A clean split 100ka, B: A clean split 200ka, C:
        population growth at 200ka followed by split 100ka, D: split 200ka
        followed by migration until 100ka, and E: split 200ka followed by
        merger 105ka and a new split 100ka. The left column shows model
        representation, the middle column shows instantaneous coalescence rates
        for pairs of lineages from the same daughter population, the right
        column shows instantaneous coalescence rates for pairs of lineages
        coming from distinct populations.  
    }
    \label{fig:toymodels}
\end{figure}

@@

\subsection{Batch effects from merged data biases inference}
\label{sec:merged-bias}
